---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# openqalyregimen

<!-- badges: start -->
[![CircleCI build status](https://circleci.com/gh/PolicyAnalysisInc/openqalyregimen.svg?style=svg)](https://app.circleci.com/pipelines/github/PolicyAnalysisInc/openqalyregimen)
[![Codecov test coverage](https://codecov.io/gh/PolicyAnalysisInc/openqalyregimen/branch/master/graph/badge.svg)](https://app.codecov.io/gh/PolicyAnalysisInc/openqalyregimen?branch=master)
<!-- badges: end -->

## Overview

The openqalyregimen package provides a declarative interface for defining treatment regimens and calculating medication costs. It handles a wide range of cases including:
- Dosing per patient, per m2 of BSA, or per kg of weight
- Oral medications with dispensing costs and oral wastage
- IV and injection medications with administration costs, vials wastage, dose rounding
- Handling of loading doses, dose escalation, and sequencing
- Correctly attributes all costs to the model cycle schedule, even when medication cycles do not align with model cycles
- Extensive debugging output to ensure transparency
- Out of the box compatability with openqaly models, but can also be used independently

## Installation

``` r
# Install the development version from GitHub
remotes::install_github("PolicyAnalysisInc/openqalyregimen")
```

## Basic Example

```{r}
library(openqalyregimen)

reg <- define_regimen(
  name = "Capecitabine",
  route = "oral",
  dose_per_admin = 2000,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 21,
  admin_days = 1:14,
  max_med_cycles = 4,
  available_strengths = c(150, 500),
  unit_cost = 2.21,
  admin_cost = 45
)

calculate_med_costs(
  reg,
  model_cycle_length = 21,
  n_cycles = 4
)
```

## Learn more

The package vignettes cover:

- Oral regimens
- IV and injection regimens
- Combinations, loading doses, and phase changes
- Use inside an `openqaly` model
