---
title: "Combination Therapies, Loading Doses, and Phase Changes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combination Therapies, Loading Doses, and Phase Changes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(openqalyregimen)
library(flextable)
```

This vignette shows how to combine multiple drugs, model loading doses, dose
escalation, and phase changes using `openqalyregimen`. It assumes familiarity
with single-drug oral and IV costing covered in `vignette("oral-regimens")`
and `vignette("iv-regimens")`.

## 1. Combination Therapies

To cost a multi-drug combination, pass multiple `med_regimen` objects to
`calculate_med_costs()`. Drugs in a combination can have different `admin_days`.
Here Drug C is given on Day 1 only, while Drug D is given on Days 1 and 8 of a
21-day cycle (a cisplatin + gemcitabine-like schedule):

```{r combo-define}
drug_day1 <- define_regimen(
  name = "IV Drug C",
  route = "iv",
  dose_per_admin = 70,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 21,
  admin_days = c(1),
  max_med_cycles = 6,
  vial_sizes = c(50, 100),
  vial_cost = c(5.21, 10.42),
  wastage_threshold = 0.1,
  admin_cost = 241
)

drug_day1_8 <- define_regimen(
  name = "IV Drug D",
  route = "iv",
  dose_per_admin = 1000,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 21,
  admin_days = c(1, 8),
  max_med_cycles = 6,
  vial_sizes = c(200, 1000, 2000),
  vial_cost = c(3.80, 15.95, 31.90),
  wastage_threshold = 0.1,
  admin_cost = 241
)
```

Setting `daily_detail = TRUE` with multiple regimens produces a wide-format
table with both drugs side by side. This makes the different schedules visible:
Drug C has cost only on Day 1, while Drug D has cost on Days 1 and 8.

The `combo_regimen_daily_detail` table has a two-row header:

- **Top row** groups columns by drug name.
- **Bottom row** shows the column type within each drug group.
- **Shared columns**: Day, Model Cycle.
- **Combined Total Cost** = sum of per-drug Total Cost columns for each day.

```{r combo-daily}
combo_daily <- calculate_med_costs(
  drug_day1, drug_day1_8,
  model_cycle_length = 21,
  n_cycles = 2,
  daily_detail = TRUE
)

as_flextable(combo_daily)
```

**Verify:** Drug C is administered on days 0 and 21 only. Drug D is
administered on days 0, 7, 21, and 28. On day 0, both drugs incur costs; on
day 7, only Drug D does.

The cycle-level `combo_regimen_costs` table aggregates daily costs into one row
per model cycle:

- **Top row** groups columns by drug name. Shared columns (Cycle, Start Day,
  End Day) and the Combined Total column are standalone.
- **Bottom row** shows the column type within each drug group: Drug Cost, Admin
  Cost, Total Cost.
- **Combined Total** column = sum of all per-drug Total Cost columns.

```{r combo-cycle}
combo_costs <- calculate_med_costs(
  drug_day1, drug_day1_8,
  model_cycle_length = 21,
  n_cycles = 6
)

as_flextable(combo_costs)
```

**Verify:** Drug C has 1 administration per cycle, Drug D has 2. Mean doses:
Drug C = 126 mg, Drug D = 1,800 mg per administration, both
distribution-averaged with BSA SD = 0.2.

Dosing diagnostics are always per-drug. Drug C:

```{r combo-dosing-c}
as_flextable(diagnose_dosing(drug_day1))
```

Drug D:

```{r combo-dosing-d}
as_flextable(diagnose_dosing(drug_day1_8))
```

## 2. Sharing Administration Costs

When multiple IV drugs are administered on the same day during a single clinic
visit, they may share infusion chair time. The `shared_admin` parameter
controls this. Without sharing (the default), each drug's `admin_cost` is
charged independently on every administration day.

With `shared_admin = TRUE`, drugs share a single administration cost on
overlapping days. The highest cost in the group is kept; the rest are zeroed
out. The daily detail makes the effect visible — on Day 1 both drugs are
administered, so only Drug C's higher admin cost (\$241) is charged. On Day 8,
only Drug D is administered, so sharing has no effect:

```{r shared-admin-daily}
shared_daily <- calculate_med_costs(
  drug_day1, drug_day1_8,
  model_cycle_length = 21,
  n_cycles = 2,
  daily_detail = TRUE,
  shared_admin = TRUE
)

as_flextable(shared_daily)
```

The `Admin (Pre-Share)` columns show the original admin costs before sharing
adjustment. At the cycle level, the savings accumulate:

```{r shared-admin-cycle}
shared_costs <- calculate_med_costs(
  drug_day1, drug_day1_8,
  model_cycle_length = 21,
  n_cycles = 6,
  shared_admin = TRUE
)

as_flextable(shared_costs)
```

**Verify:** On Day 1, both drugs have \$241 admin cost. With sharing, Drug C
wins (first by input order on ties) and Drug D's admin is zeroed. On Day 8,
only Drug D is present so its \$241 is charged normally. Per cycle: \$241 saved
vs. no sharing.

You can also use explicit groups when only some drugs share admin costs:

```r
calculate_med_costs(
  drug_a, drug_b, drug_c,
  shared_admin = list(c("Drug A", "Drug B"))  # Drug C is independent
)
```

## 3. Costing an IV + Oral Combination

IV and oral drugs combine seamlessly. Here an IV drug is given on Day 1 and an
oral drug is taken on Days 1--14 of a 21-day cycle (a CAPOX-like schedule). If
the clinical regimen is twice daily, the table or direct definition should
encode the total daily dose in `dose_per_admin`; `admin_days` only identifies
which days the drug is taken:

```{r combo-mixed}
iv_drug <- define_regimen(
  name = "IV Drug E",
  route = "iv",
  dose_per_admin = 130,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 21,
  admin_days = c(1),
  max_med_cycles = 8,
  vial_sizes = c(50, 100, 200),
  vial_cost = c(12.32, 24.64, 49.28),
  wastage_threshold = 0.1,
  admin_cost = 241
)

oral_drug <- define_regimen(
  name = "Oral Drug F",
  route = "oral",
  dose_per_admin = 2000,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  med_cycle_length = 21,
  admin_days = 1:14,
  max_med_cycles = 8,
  available_strengths = c(150, 500),
  unit_cost = 2.21,
  admin_cost = 0
)

combo_mixed <- calculate_med_costs(
  iv_drug, oral_drug,
  model_cycle_length = 21,
  n_cycles = 8
)

as_flextable(combo_mixed)
```

**Verify:** The combination cost sums both components. IV Drug E mean dose =
130 x 1.8 = 234 mg (distribution-averaged). Oral Drug F daily dose = 2,000 x
1.8 = 3,600 mg on each treatment day, with 14 treatment days per cycle.

Dosing diagnostic for the IV component:

```{r combo-mixed-dosing}
as_flextable(diagnose_dosing(iv_drug))
```

The daily detail view shows both drugs side by side. IV drugs get vial-specific
columns (mg Vials, mg Waste, Vials); oral drugs get tablet/dispensing columns
(Disp?, Tabs, etc.):

```{r combo-mixed-daily}
combo_mixed_daily <- calculate_med_costs(
  iv_drug, oral_drug,
  model_cycle_length = 21,
  n_cycles = 2,
  daily_detail = TRUE
)

as_flextable(combo_mixed_daily)
```

## 4. Modeling a Loading Dose

Some drugs use a higher first dose (loading) followed by a lower maintenance
dose. Model this as two separate regimen objects with `start_day` offsets.

The key formula for phase sequencing: a phase runs from `start_day` to
`start_day + max_med_cycles * med_cycle_length`. The next phase's `start_day`
should equal the previous phase's end.

Here a drug has an 8 mg/kg loading dose for 1 cycle, then 6 mg/kg maintenance
(a trastuzumab-like schedule). The loading phase occupies days 0--20
(`start_day = 0, max_med_cycles = 1, med_cycle_length = 21`), and the
maintenance phase begins on day 21:

```{r loading}
loading <- define_regimen(
  name = "Drug G (loading)",
  route = "iv",
  dose_per_admin = 8,
  dose_basis = "weight",
  patient_weight = 70,
  patient_weight_sd = 12,
  med_cycle_length = 21,
  admin_days = 1,
  max_med_cycles = 1,
  start_day = 0,
  vial_sizes = c(150, 420),
  vial_cost = c(407.40, 1184),
  wastage_threshold = 0.1
)

maintenance <- define_regimen(
  name = "Drug G (maintenance)",
  route = "iv",
  dose_per_admin = 6,
  dose_basis = "weight",
  patient_weight = 70,
  patient_weight_sd = 12,
  med_cycle_length = 21,
  admin_days = 1,
  max_med_cycles = Inf,
  start_day = 21,
  vial_sizes = c(150, 420),
  vial_cost = c(407.40, 1184),
  wastage_threshold = 0.1
)

loading_costs <- calculate_med_costs(
  loading, maintenance,
  model_cycle_length = 21,
  n_cycles = 26
)

as_flextable(loading_costs)
```

**Verify:** Loading dose = 8 x 70 = 560 mg incurs a higher cost in cycle 1.
Maintenance dose = 6 x 70 = 420 mg runs from cycle 2 onward. Both phases use
distribution-averaged costs with `patient_weight_sd = 12`.

Dosing diagnostics confirm different vial selections for the loading phase:

```{r loading-dosing-load}
as_flextable(diagnose_dosing(loading))
```

And the maintenance phase:

```{r loading-dosing-maint}
as_flextable(diagnose_dosing(maintenance))
```

## 5. Modeling Dose Escalation

Multi-week dose ramp-ups are modeled by chaining multiple regimen objects with
sequential `start_day` values. The formula for the next start day:
`next_start_day = current_start_day + max_med_cycles * med_cycle_length`.

Here a drug ramps through 20/50/100/200 mg/day over 4 weeks before reaching
400 mg/day maintenance (a venetoclax-like schedule):

```{r escalation}
strengths <- c(10, 50, 100)
cost_tab  <- 42.76

wk1 <- define_regimen(
  name = "Escalation Wk1", route = "oral",
  dose_per_admin = 20, med_cycle_length = 7,
  admin_days = 1:7, max_med_cycles = 1, start_day = 0,
  available_strengths = strengths, unit_cost = cost_tab
)

wk2 <- define_regimen(
  name = "Escalation Wk2", route = "oral",
  dose_per_admin = 50, med_cycle_length = 7,
  admin_days = 1:7, max_med_cycles = 1, start_day = 7,
  available_strengths = strengths, unit_cost = cost_tab
)

wk3 <- define_regimen(
  name = "Escalation Wk3", route = "oral",
  dose_per_admin = 100, med_cycle_length = 7,
  admin_days = 1:7, max_med_cycles = 1, start_day = 14,
  available_strengths = strengths, unit_cost = cost_tab
)

wk4 <- define_regimen(
  name = "Escalation Wk4", route = "oral",
  dose_per_admin = 200, med_cycle_length = 7,
  admin_days = 1:7, max_med_cycles = 1, start_day = 21,
  available_strengths = strengths, unit_cost = cost_tab
)

maint <- define_regimen(
  name = "Maintenance", route = "oral",
  dose_per_admin = 400, med_cycle_length = 28,
  admin_days = 1:28, max_med_cycles = Inf, start_day = 28,
  available_strengths = strengths, unit_cost = cost_tab
)

esc_costs <- calculate_med_costs(
  wk1, wk2, wk3, wk4, maint,
  model_cycle_length = 7,
  n_cycles = 52
)

as_flextable(esc_costs)
```

**Verify:** Weeks 1--4 each show escalating costs reflecting 20, 50, 100, and
200 mg daily doses. From week 5 onward, the maintenance phase (400 mg = 4 x
100 mg tablets) dominates.
