---
title: "Costing IV and Injection Regimens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Costing IV and Injection Regimens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(openqalyregimen)
library(flextable)
```

This vignette shows how to cost intravenous (IV), subcutaneous (SC), and
intramuscular (IM) drug regimens using `openqalyregimen`. Each section walks
through a specific use case, introducing functions and parameters as they become
relevant.

For oral drugs, see `vignette("oral-regimens")`. For multi-drug combinations,
loading doses, and phase changes, see `vignette("combinations-and-phases")`.

## 1. Costing a Flat-Dose IV Drug

The simplest IV regimen is a flat dose with a single vial size. Here a drug is
dosed at 200 mg using 100 mg vials (a pembrolizumab-like schedule).

Start by defining the regimen with `define_regimen()`, then use
`summarise_regimen()` to inspect the vial breakdown for a single
administration:

```{r iv-flat-define}
reg_flat <- define_regimen(
  name = "IV Drug A 200mg Q3W",
  route = "iv",
  dose_per_admin = 200,
  dose_basis = "flat",
  med_cycle_length = 21,
  admin_days = 1,
  max_med_cycles = 35,
  vial_sizes = 100,
  vial_cost = 2630,
  admin_cost = 241
)

summarise_regimen(reg_flat)
```

Now calculate cycle-by-cycle costs with `calculate_med_costs()`. The cost table
has one row per model cycle with the following columns:

- **Cycle** -- Sequential model cycle number.
- **Start Day** / **End Day** -- Model cycle boundaries.
- **N Admins** -- Number of IV administrations in this model cycle.
- **Drug Cost** -- Total drug acquisition cost (N Admins x per-admin drug cost).
- **Admin Cost** -- Total administration cost (N Admins x per-admin admin cost).
  The `admin_cost` parameter covers infusion chair time, nursing, etc.
- **Total Cost** -- Drug Cost + Admin Cost.

```{r iv-flat-costs}
costs_flat <- calculate_med_costs(
  reg_flat,
  model_cycle_length = 21,
  n_cycles = 35
)

as_flextable(costs_flat)
```

**Verify:** Each administration uses 2 x 100 mg vials = 200 mg with zero
wastage. Drug Cost per admin = 2 x \$2,630 = \$5,260. Admin Cost = \$241 per
admin. Total Cost = \$5,501 per cycle.

Setting `daily_detail = TRUE` produces a row-per-day breakdown showing vial
usage on each administration day. This is useful for verifying exactly which
days incur costs.

```{r iv-flat-daily}
daily_iv <- calculate_med_costs(
  reg_flat,
  model_cycle_length = 21,
  n_cycles = 2,
  daily_detail = TRUE
)
as_flextable(daily_iv)
```

The `regimen_daily_detail` table (IV variant) has a two-row header with groups:

- **Medication Cycle** group: *Med Cycle*, *Cycle Day*.
- **Dosing Detail** group: *Admin?*, *Doses*, *mg Taken*, *mg Vials*
  (total drug content in opened vials), *mg Waste* (drug discarded), *Vials*
  (number opened).
- **Costs** group: *Drug Cost*, *Admin Cost*, *Total Cost*.

On administration days, mg Vials = mg Taken + mg Waste. On non-administration
days, all dosing and cost columns are zero.

## 2. Costing with Multiple Vial Sizes

When multiple vial sizes are available, the algorithm selects vials using the
following priority hierarchy, consistent with the Australian PBS vial
combination algorithm and CMS billing guidance:

1. **Exact fill.** If any combination of available vials sums to exactly the
   prescribed dose, that combination is used. Among multiple exact-fill options,
   the cheapest is selected. This eliminates avoidable drug waste -- drug that
   would be opened but not administered.

2. **Under-fill within tolerance.** When no exact fill exists and a
   `wastage_threshold` is specified, the algorithm searches for the closest
   achievable dose below the prescribed dose. If the shortfall would only be
   covered by opening a vial that is predominantly wasted (fraction used <=
   threshold), and the under-fill is cheaper than the overshoot option, the
   under-fill is selected. This implements dose banding: the patient receives a
   slightly reduced dose to avoid opening a barely-used vial.

3. **Overshoot.** When neither exact fill nor an acceptable, cheaper under-fill
   is available, the algorithm finds the cheapest combination of vials whose
   total content exceeds the prescribed dose. The patient receives their full
   prescribed dose; excess drug remains in the vial. Among equal-cost options,
   the combination with less total drug content (less waste) is preferred.

The selection within each tier is solved via dynamic programming. The decision
logic is summarised below:

```text
                  Can any combination of
                vials sum to exactly the dose?
                     /              \
                   YES               NO
                    |                 |
          Use cheapest         Is wastage_threshold > 0?
          exact fill.              /          \
          Waste = 0.             YES           NO
                                  |             |
                    Is there an achievable       |
                    dose below prescribed        |
                    where the shortfall          |
                    fraction <= threshold?        |
                        /          \             |
                      YES           NO           |
                       |             \          /
              Is under-fill       Use cheapest
              cheaper than        combination that
              overshoot?          covers the dose.
               /       \
             YES        NO
              |          |
        Use the      Use cheapest
        under-fill.  combination that
        Waste = 0.   covers the dose.
```

Here a drug is dosed at 240 mg flat with vial sizes 40, 100, 120, and 240 mg (a
nivolumab-like schedule):

```{r iv-multi-vial}
reg_multi <- define_regimen(
  name = "IV Drug B 240mg Q2W",
  route = "iv",
  dose_per_admin = 240,
  dose_basis = "flat",
  med_cycle_length = 14,
  admin_days = 1,
  vial_sizes = c(40, 100, 120, 240),
  vial_cost = c(439, 1097, 1317, 2633),
  admin_cost = 0
)

summarise_regimen(reg_multi)
```

**Verify:** 240 mg is filled exactly with zero wastage at \$2,633. Multiple
combinations achieve this — the algorithm selects 2 x 100 mg + 1 x 40 mg,
which ties in cost with 1 x 240 mg.

Now consider a dose where no combination of available vial sizes sums exactly
to the prescribed amount. At 250 mg, the algorithm cannot achieve an exact fill
with 40/100/120/240 mg vials, so it falls back to the overshoot tier (tier 3)
and selects the cheapest covering combination — 1 x 100 mg + 4 x 40 mg =
260 mg — delivering the full 250 mg dose with 10 mg of waste:

```{r iv-multi-vial-residual}
reg_residual <- define_regimen(
  name = "IV Drug B 250mg",
  route = "iv",
  dose_per_admin = 250,
  dose_basis = "flat",
  med_cycle_length = 14,
  admin_days = 1,
  vial_sizes = c(40, 100, 120, 240),
  vial_cost = c(439, 1097, 1317, 2633),
  admin_cost = 0
)

summarise_regimen(reg_residual)
```

**Verify:** No combination of 40/100/120/240 mg vials sums to exactly 250 mg,
so the algorithm overshoots to 260 mg (1 x 100 mg + 4 x 40 mg) with 10 mg
waste. Drug Cost = \$1,097 + 4 x \$439 = \$2,853.

### Wastage threshold and dose banding

The `wastage_threshold` parameter (between 0.0 and 1.0) activates the
under-fill tier (tier 2) of the vial selection hierarchy. When no exact fill
exists, the algorithm looks for the closest achievable dose below the prescribed
dose. If the shortfall from the prescribed dose would require opening a vial
where the fraction used is at or below the threshold, and the under-fill is
cheaper than the overshoot alternative, the algorithm selects the under-fill and
the patient receives a slightly reduced dose.

Here a drug is dosed at 850 mg with 200 mg and 500 mg vials. Without a wastage
threshold, the algorithm overshoots: 1 x 500 mg + 2 x 200 mg = 900 mg,
delivering the full dose with 50 mg waste:

```{r iv-wastage}
reg_wastage <- define_regimen(
  name = "IV Drug C 850mg",
  route = "iv",
  dose_per_admin = 850,
  dose_basis = "flat",
  med_cycle_length = 28,
  admin_days = c(1, 15),
  vial_sizes = c(200, 500),
  vial_cost = c(1500, 3500),
  wastage_threshold = 0,
  admin_cost = 0
)

summarise_regimen(reg_wastage)
```

With `wastage_threshold = 0.25`, the algorithm finds that the 50 mg shortfall
would use only 50/200 = 25% of a 200 mg vial, which is at the threshold. It
selects the under-fill instead: 4 x 200 mg = 800 mg with zero waste at \$6,000:

```{r iv-wastage-zero}
reg_wastage_thresh <- define_regimen(
  name = "IV Drug C 850mg (threshold 0.25)",
  route = "iv",
  dose_per_admin = 850,
  dose_basis = "flat",
  med_cycle_length = 28,
  admin_days = c(1, 15),
  vial_sizes = c(200, 500),
  vial_cost = c(1500, 3500),
  wastage_threshold = 0.25,
  admin_cost = 0
)

summarise_regimen(reg_wastage_thresh)
```

**Verify:** Without threshold: 1 x 500 mg + 2 x 200 mg = 900 mg, cost =
\$3,500 + 2 x \$1,500 = \$6,500, waste = 50 mg. With threshold 0.25: the
marginal 200 mg vial is skipped (50/200 = 0.25 <= 0.25), so 4 x 200 mg =
800 mg, cost = 4 x \$1,500 = \$6,000, waste = 0 mg.

## 3. Costing a Weight-Based IV Drug

For weight-based dosing, specify `dose_basis = "weight"` and `patient_weight`.
This produces a cost for a single patient at exactly that weight. Here a drug is
dosed at 3 mg/kg for a 70 kg patient with multiple vial sizes (a nivolumab-like
schedule).

First, the vial breakdown for a single administration:

```{r iv-weight-define}
reg_weight <- define_regimen(
  name = "IV Drug B 3mg/kg Q2W",
  route = "iv",
  dose_per_admin = 3,
  dose_basis = "weight",
  patient_weight = 70,
  med_cycle_length = 14,
  admin_days = 1,
  vial_sizes = c(40, 100, 120, 240),
  vial_cost = c(439, 1097, 1317, 2633),
  admin_cost = 0
)

summarise_regimen(reg_weight)
```

And the cycle-by-cycle costs:

```{r iv-weight-costs}
costs_weight <- calculate_med_costs(
  reg_weight,
  model_cycle_length = 28,
  n_cycles = 26,
  subcycle_precision = TRUE
)

as_flextable(costs_weight)
```

**Verify:** Dose = 3 x 70 = 210 mg. The optimal packing finds the cheapest
vial combination covering 210 mg. Check `summarise_regimen()` output for the
exact vial selection and cost.

### Distribution-averaged costs

In cohort models, we typically want the expected cost across a
population of patients, not just the cost for one specific patient. Adding
`patient_weight_sd` enables distribution-averaged costing.

The package models the patient parameter with a lognormal distribution,
identifies cost breakpoints where the vial combination changes, and computes a
probability-weighted average cost. This matters because the cost function is a
step function: the expected cost across weights does not equal the cost at the
expected weight.

```{r iv-dist-define}
reg_dist <- define_regimen(
  name = "IV Drug B 3mg/kg Q2W",
  route = "iv",
  dose_per_admin = 3,
  dose_basis = "weight",
  patient_weight = 70,
  patient_weight_sd = 12,
  med_cycle_length = 14,
  admin_days = 1,
  vial_sizes = c(40, 100, 120, 240),
  vial_cost = c(439, 1097, 1317, 2633),
  admin_cost = 0
)

summarise_regimen(reg_dist)
```

The cycle-by-cycle costs now reflect the distribution-averaged per-admin cost:

```{r iv-dist-costs}
costs_dist <- calculate_med_costs(
  reg_dist,
  model_cycle_length = 28,
  n_cycles = 26,
  subcycle_precision = TRUE
)

as_flextable(costs_dist)
```

**Verify:** Compare the per-cycle drug cost here with the point-estimate cost
from the fixed-weight example above. The distribution-averaged cost will differ
because patients above and below the mean weight may fall into different vial
combinations with different costs.

### Inspecting dose bins with `diagnose_dosing()`

The `diagnose_dosing()` function produces a detailed breakdown of how patient
parameter variation maps to dose bins, vial combinations, probabilities, and
costs. This is the primary diagnostic table for verifying distribution-averaged
costs.

The `dosing_detail` table has one row per dose bin with the following columns:

- **Bin** -- Sequential bin number. Each bin represents a range of patient
  values that produce the same vial combination.
- **Prob.** -- Probability that a randomly drawn patient falls in this bin.
  Values across all bins sum to approximately 1.0.
- **Patient Range (kg or m^2^)** -- Range of patient parameter values that
  produce the same vial combination.
- **Dose Rate (mg/kg or mg/m^2^)** -- The prescribed dose rate. This is
  constant across all bins.
- **Dose Range (mg)** -- The actual dose range for patients in this bin,
  calculated as Patient Range x Dose Rate.
- **Vial Combination** -- Which vials are used. "NxSIZEmg" denotes fully
  consumed vials. "+ open NxSIZEmg" denotes the final partially-used vial.
- **N Vials** -- Total number of vials opened for one administration.
- **Total Vial (mg)** -- Total drug content across all opened vials. This is
  always >= the dose administered.
- **Waste (mg)** -- Drug discarded from partially-used vials
  (Total Vial mg - dose administered).
- **Drug Cost** -- Acquisition cost for one administration in this bin.
- **Weighted Cost** -- Drug Cost x Probability. The sum of Weighted Cost across
  all bins equals the distribution-averaged cost.

The **footer** shows the expected (distribution-averaged) cost, drug name, dose
basis, and distribution parameters (mean +/- SD).

```{r iv-dosing-diag}
dosing_diag <- diagnose_dosing(reg_dist)
as_flextable(dosing_diag)
```

**Verify:** Find the bin containing the mean patient value (70 kg) and confirm
it matches the point-estimate cost from the fixed-weight example. Check that
probabilities sum to ~1.0. Confirm that the sum of the Weighted Cost column
equals the Expected Cost shown in the footer.

## 4. Costing a BSA-Based IV Drug

BSA-based IV dosing works identically to weight-based dosing, but the patient
parameter is BSA in m^2^ instead of weight in kg. Here a drug is dosed at
375 mg/m^2^ with BSA = 1.8 m^2^, BSA SD = 0.2, vial sizes 100 and 500 mg,
and a 10% wastage threshold (a rituximab-like schedule).

The cost table:

```{r iv-bsa-costs}
reg_bsa <- define_regimen(
  name = "IV Drug D 375mg/m2",
  route = "iv",
  dose_per_admin = 375,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 7,
  admin_days = 1,
  max_med_cycles = 4,
  vial_sizes = c(100, 500),
  vial_cost = c(174.63, 873.15),
  wastage_threshold = 0.1,
  admin_cost = 0
)

costs_bsa <- calculate_med_costs(
  reg_bsa,
  model_cycle_length = 7,
  n_cycles = 4
)

as_flextable(costs_bsa)
```

The dosing diagnostic table has the same structure as for weight-based dosing,
but Patient Range shows m^2^ instead of kg:

```{r iv-bsa-dosing}
as_flextable(diagnose_dosing(reg_bsa))
```

**Verify:** Mean dose = 375 x 1.8 = 675 mg. With `patient_bsa_sd = 0.2`,
costs are distribution-averaged across dose bins. Confirm only 4 doses are
given across the 4-week period (max_med_cycles = 4).

## 5. Costing SC and IM Drugs

Setting `route = "sc"` (subcutaneous) or `route = "im"` (intramuscular) uses
the same vial-based costing logic as IV. All three are `VIAL_ROUTES` in the
package.

Here a drug is dosed at 1.3 mg/m^2^ SC with 1 mg and 2 mg vials on days 1, 4,
8, and 11 of a 21-day cycle (a bortezomib-like schedule). Using two small vial
sizes means the optimal vial combination changes across the BSA distribution,
producing multiple dose bins. First, the vial breakdown:

```{r sc-define}
reg_sc <- define_regimen(
  name = "SC Drug E 1.3mg/m2",
  route = "sc",
  dose_per_admin = 1.3,
  dose_basis = "bsa",
  patient_bsa = 1.8,
  patient_bsa_sd = 0.2,
  med_cycle_length = 21,
  admin_days = c(1, 4, 8, 11),
  vial_sizes = c(1, 2),
  vial_cost = c(218, 436),
  wastage_threshold = 0.1
)

summarise_regimen(reg_sc)
```

The cycle-by-cycle costs:

```{r sc-costs}
costs_sc <- calculate_med_costs(
  reg_sc,
  model_cycle_length = 21,
  n_cycles = 8,
  subcycle_precision = TRUE
)

as_flextable(costs_sc)
```

**Verify:** Mean dose = 1.3 x 1.8 = 2.34 mg. With 1 mg and 2 mg vials, the
vial combination varies across the BSA distribution — smaller patients need
fewer vials than larger patients. The SC route is costed identically to IV.

The dosing diagnostic for the SC drug shows multiple dose bins:

```{r sc-dosing}
as_flextable(diagnose_dosing(reg_sc))
```

## 6. Assigning Regimen Costs to Model Cycles

When the medication cycle length (`med_cycle_length`) differs from the model
cycle length (`model_cycle_length`), administration events don't land evenly
across model cycles. The package offers two ways to handle this:

- **Default (analytic):** An approximation that counts events per model cycle.
  Exact when cycles align; may accumulate small rounding errors when they
  don't.
- **`subcycle_precision = TRUE`:** Builds a full day-level timeline for exact
  event counting. Always precise.

Here is a 21-day medication cycle with IV dosing on days 1, 4, 8, and
11, mapped onto 30-day model cycles. Because the administration days
are sparse and unevenly spaced, different model cycles capture
different numbers of doses — making the contrast between the two
methods clear:

```{r subcycle-precision-demo}
reg_demo <- define_regimen(
  name = "IV Drug (demo)",
  route = "iv",
  dose_per_admin = 200,
  vial_sizes = 100,
  vial_cost = 2630,
  med_cycle_length = 21,
  admin_days = c(1, 4, 8, 11),
  admin_cost = 241
)

# Analytic approximation (default)
costs_approx <- calculate_med_costs(
  reg_demo,
  model_cycle_length = 30,
  n_cycles = 12
)

as_flextable(costs_approx)
```

```{r subcycle-precision-exact}
# Exact subcycle precision
costs_exact <- calculate_med_costs(
  reg_demo,
  model_cycle_length = 30,
  n_cycles = 12,
  subcycle_precision = TRUE
)

as_flextable(costs_exact)
```

**Verify:** Both tables span 360 days (12 x 30). The analytic method
assigns a uniform fractional administration count per cycle. Subcycle
precision shows exact integer counts that vary across cycles depending
on which administration days fall within each model cycle window.
