---
title: "Using openqalyregimen in an openqaly Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using openqalyregimen in an openqaly Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(openqaly)
library(openqalyregimen)
library(dplyr)
library(tibble)
```

This vignette shows how to use `openqalyregimen` inside a working
`openqaly` model. Medication and administration costing can become complicated
when treatment regimens include loading doses, maintenance dosing, different
routes of administration, and route-specific costing rules.

`openqalyregimen` provides a declarative way to define those regimens and turn
them into cost vectors. `openqaly` can then use those vectors as ordinary model
inputs.

This vignette shows two ways to define the same treatment logic:

- direct regimen definitions with `define_regimen()`
- table-based regimen definitions with `define_regimen_table()`

Both approaches compare:

- `oral_mono`, an oral capecitabine monotherapy
- `iv_combo`, an IV trastuzumab plus pertuzumab regimen

The trastuzumab strategy includes a loading dose followed by maintenance
dosing. The model keeps those phases together as one medication-level cost
stream while still separating medication and administration line items.

The model uses the standard `openqaly` pattern. Values represent line items.
Summaries aggregate those values into outputs such as total cost, total QALYs,
and total life-years.

## Shared model structure

Both sections use the same Markov model:

- states: `alive`, `dead`
- cycle length: 21 days
- time horizon: 8 cycles

Patient-level dosing inputs stay outside the regimen tables:

- `patient_bsa`
- `patient_weight`

```{r base-model}
base_model <- define_model("markov") |>
  set_settings(
    n_cycles = 8,
    cycle_length = 21,
    cycle_length_unit = "days"
  ) |>
  add_strategy("oral_mono", "Oral Monotherapy") |>
  add_strategy("iv_combo", "IV Combination") |>
  add_state("alive", initial_prob = 1) |>
  add_state("dead", initial_prob = 0) |>
  add_transition("alive", "dead", 0.08) |>
  add_transition("alive", "alive", C) |>
  add_transition("dead", "dead", 1) |>
  add_variable("patient_bsa", 1.8) |>
  add_variable("patient_weight", 70)
```

## 1. Direct regimen definitions with `define_regimen()`

This version defines each regimen directly inside model variables. Capecitabine
appears only in `oral_mono`. The `iv_combo` strategy contains trastuzumab
loading, trastuzumab maintenance, and pertuzumab. The trastuzumab loading and
maintenance vectors are combined before they are added to the model as values,
so they remain one medication-level cost stream in the results.

```{r direct-model}
model_direct <- base_model |>
  add_variable(
    "capecitabine_regimen",
    define_regimen(
      name = "Capecitabine",
      route = "oral",
      dose_per_admin = 2000,
      dose_basis = "bsa",
      patient_bsa = patient_bsa,
      med_cycle_length = cycle_length_days,
      admin_days = 1:14,
      max_med_cycles = max(cycle),
      available_strengths = c(150, 500),
      unit_cost = 2.21,
      admin_cost = 45
    ),
    strategy = "oral_mono"
  ) |>
  add_variable(
    "trastuzumab_loading_regimen",
    define_regimen(
      name = "Trastuzumab loading",
      route = "iv",
      dose_per_admin = 8,
      dose_basis = "weight",
      patient_weight = patient_weight,
      med_cycle_length = cycle_length_days,
      admin_days = 1,
      max_med_cycles = 1,
      start_day = 0,
      vial_sizes = c(150, 420),
      vial_cost = c(407.40, 1184),
      wastage_threshold = 0.1,
      admin_cost = 241
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "trastuzumab_maintenance_regimen",
    define_regimen(
      name = "Trastuzumab maintenance",
      route = "iv",
      dose_per_admin = 6,
      dose_basis = "weight",
      patient_weight = patient_weight,
      med_cycle_length = cycle_length_days,
      admin_days = 1,
      max_med_cycles = max(cycle) - 1,
      start_day = cycle_length_days,
      vial_sizes = c(150, 420),
      vial_cost = c(407.40, 1184),
      wastage_threshold = 0.1,
      admin_cost = 241
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "pertuzumab_regimen",
    define_regimen(
      name = "Pertuzumab",
      route = "iv",
      dose_per_admin = 420,
      dose_basis = "flat",
      med_cycle_length = cycle_length_days,
      admin_days = 1,
      max_med_cycles = max(cycle),
      vial_sizes = 420,
      vial_cost = 2750,
      admin_cost = 241
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_capecitabine_med",
    calculate_regimen_cost(
      capecitabine_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "oral_mono"
  ) |>
  add_variable(
    "c_capecitabine_admin",
    calculate_regimen_cost(
      capecitabine_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "oral_mono"
  ) |>
  add_variable("c_trastuzumab_med", 0, strategy = "oral_mono") |>
  add_variable("c_trastuzumab_admin", 0, strategy = "oral_mono") |>
  add_variable("c_pertuzumab_med", 0, strategy = "oral_mono") |>
  add_variable("c_pertuzumab_admin", 0, strategy = "oral_mono") |>
  add_variable("c_capecitabine_med", 0, strategy = "iv_combo") |>
  add_variable("c_capecitabine_admin", 0, strategy = "iv_combo") |>
  add_variable(
    "c_trastuzumab_med",
    calculate_regimen_cost(
      trastuzumab_loading_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ) +
      calculate_regimen_cost(
        trastuzumab_maintenance_regimen,
        type = "medication",
        model_cycle_length = cycle_length_days,
        n_cycles = max(cycle)
      ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_trastuzumab_admin",
    calculate_regimen_cost(
      trastuzumab_loading_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ) +
      calculate_regimen_cost(
        trastuzumab_maintenance_regimen,
        type = "administration",
        model_cycle_length = cycle_length_days,
        n_cycles = max(cycle)
      ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_pertuzumab_med",
    calculate_regimen_cost(
      pertuzumab_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_pertuzumab_admin",
    calculate_regimen_cost(
      pertuzumab_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_value("capecitabine_med_cost", c_capecitabine_med, state = "alive", type = "cost") |>
  add_value("capecitabine_admin_cost", c_capecitabine_admin, state = "alive", type = "cost") |>
  add_value("trastuzumab_med_cost", c_trastuzumab_med, state = "alive", type = "cost") |>
  add_value("trastuzumab_admin_cost", c_trastuzumab_admin, state = "alive", type = "cost") |>
  add_value("pertuzumab_med_cost", c_pertuzumab_med, state = "alive", type = "cost") |>
  add_value("pertuzumab_admin_cost", c_pertuzumab_admin, state = "alive", type = "cost") |>
  add_value("qaly", 0.8 * cycle_length_years, state = "alive", type = "outcome") |>
  add_value("ly", cycle_length_years, state = "alive", type = "outcome") |>
  add_summary(
    "total_cost",
    "capecitabine_med_cost,capecitabine_admin_cost,trastuzumab_med_cost,trastuzumab_admin_cost,pertuzumab_med_cost,pertuzumab_admin_cost",
    type = "cost"
  ) |>
  add_summary("total_qalys", "qaly", type = "outcome") |>
  add_summary("total_lys", "ly", type = "outcome")

results_direct <- run_model(model_direct)
direct_cost_breakdown <- get_summaries(
  results_direct,
  summaries = "total_cost",
  value_type = "cost",
  discounted = TRUE,
  use_display_names = FALSE
)
```

The model can now be run. The output below shows the `total_cost` summary
broken down into its contributing value components.

```{r direct-cost-breakdown}
direct_cost_breakdown |>
  select(strategy, group, summary, value, amount)
```

This is where the medication and administration line items appear. They are
values inside the `total_cost` summary, not separate summaries.

## 2. Table-based regimen definitions with `define_regimen_table()`

This version defines the same treatment logic from one regimen table. Each row
describes one regimen component. The table is then filtered to select the rows
needed for each regimen object. Patient-level inputs such as BSA and weight
remain in the model and are passed separately. The trastuzumab rows include a
loading row and a maintenance row, and those rows are passed together into one
regimen object so they remain one medication-level stream in the model outputs.

```{r regimen-tables}
regimen_tbl <- tribble(
  ~component_group,          ~name,                      ~route, ~dose, ~dose_basis, ~med_cycle_length, ~admin_days, ~max_cycles, ~start_day, ~available_strengths, ~unit_cost, ~vial_sizes, ~vial_cost,      ~wastage_threshold, ~admin_cost,
  "oral_mono",               "Capecitabine",             "oral",  2000, "bsa",                       21, "1-14",               8,         NA, "150,500",                     2.21, "",          "",                           NA,          45,
  "iv_combo_trastuzumab",    "Trastuzumab loading",      "iv",       8, "weight",                    21, "1",                  1,          0, "",                             NA, "150,420",   "407.40,1184",                0.1,         241,
  "iv_combo_trastuzumab",    "Trastuzumab maintenance",  "iv",       6, "weight",                    21, "1",                  7,         21, "",                             NA, "150,420",   "407.40,1184",                0.1,         241,
  "iv_combo_pertuzumab",     "Pertuzumab",               "iv",     420, "flat",                      21, "1",                  8,         NA, "",                             NA, "420",       "2750",                       NA,         241
)

regimen_tbl
```

```{r tabular-model}
model_tabular <- base_model |>
  add_table("regimen_tbl", regimen_tbl) |>
  add_variable(
    "capecitabine_regimen",
    define_regimen_table(
      regimen_tbl |>
        filter(component_group == "oral_mono"),
      patient_bsa = patient_bsa
    ),
    strategy = "oral_mono"
  ) |>
  add_variable(
    "trastuzumab_regimen",
    define_regimen_table(
      regimen_tbl |>
        filter(component_group == "iv_combo_trastuzumab"),
      patient_weight = patient_weight
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "pertuzumab_regimen",
    define_regimen_table(
      regimen_tbl |>
        filter(component_group == "iv_combo_pertuzumab")
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_capecitabine_med",
    calculate_regimen_cost(
      capecitabine_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "oral_mono"
  ) |>
  add_variable(
    "c_capecitabine_admin",
    calculate_regimen_cost(
      capecitabine_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "oral_mono"
  ) |>
  add_variable("c_trastuzumab_med", 0, strategy = "oral_mono") |>
  add_variable("c_trastuzumab_admin", 0, strategy = "oral_mono") |>
  add_variable("c_pertuzumab_med", 0, strategy = "oral_mono") |>
  add_variable("c_pertuzumab_admin", 0, strategy = "oral_mono") |>
  add_variable("c_capecitabine_med", 0, strategy = "iv_combo") |>
  add_variable("c_capecitabine_admin", 0, strategy = "iv_combo") |>
  add_variable(
    "c_trastuzumab_med",
    calculate_regimen_cost(
      trastuzumab_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_trastuzumab_admin",
    calculate_regimen_cost(
      trastuzumab_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_pertuzumab_med",
    calculate_regimen_cost(
      pertuzumab_regimen,
      type = "medication",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_variable(
    "c_pertuzumab_admin",
    calculate_regimen_cost(
      pertuzumab_regimen,
      type = "administration",
      model_cycle_length = cycle_length_days,
      n_cycles = max(cycle)
    ),
    strategy = "iv_combo"
  ) |>
  add_value("capecitabine_med_cost", c_capecitabine_med, state = "alive", type = "cost") |>
  add_value("capecitabine_admin_cost", c_capecitabine_admin, state = "alive", type = "cost") |>
  add_value("trastuzumab_med_cost", c_trastuzumab_med, state = "alive", type = "cost") |>
  add_value("trastuzumab_admin_cost", c_trastuzumab_admin, state = "alive", type = "cost") |>
  add_value("pertuzumab_med_cost", c_pertuzumab_med, state = "alive", type = "cost") |>
  add_value("pertuzumab_admin_cost", c_pertuzumab_admin, state = "alive", type = "cost") |>
  add_value("qaly", 0.8 * cycle_length_years, state = "alive", type = "outcome") |>
  add_value("ly", cycle_length_years, state = "alive", type = "outcome") |>
  add_summary(
    "total_cost",
    "capecitabine_med_cost,capecitabine_admin_cost,trastuzumab_med_cost,trastuzumab_admin_cost,pertuzumab_med_cost,pertuzumab_admin_cost",
    type = "cost"
  ) |>
  add_summary("total_qalys", "qaly", type = "outcome") |>
  add_summary("total_lys", "ly", type = "outcome")

results_tabular <- run_model(model_tabular)
tabular_cost_breakdown <- get_summaries(
  results_tabular,
  summaries = "total_cost",
  value_type = "cost",
  discounted = TRUE,
  use_display_names = FALSE
)
```

```{r tabular-cost-breakdown}
tabular_cost_breakdown |>
  select(strategy, group, summary, value, amount)
```

The table provides a compact declarative input format, but the model still
uses the result in the same way. The regimen object is defined once, cost
vectors are calculated from it, and those vectors are then used as ordinary
model values.

This is the core integration pattern:

1. define regimen objects inside ordinary `openqaly` variables
2. calculate medication/admin cost vectors as variables
3. add those vectors to the model as line-item values
4. aggregate them with summaries such as `total_cost`, `total_qalys`, and
   `total_lys`
