---
title: "Costing Oral Regimens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Costing Oral Regimens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(openqalyregimen)
library(flextable)
```

This vignette shows how to cost oral drug regimens using `openqalyregimen`. Each
section walks through a specific use case, introducing functions and parameters
as they become relevant.

For IV/SC/IM drugs, see `vignette("iv-regimens")`. For multi-drug combinations,
loading doses, and phase changes, see `vignette("combinations-and-phases")`.

## 1. Costing a Daily Oral Drug

The simplest oral regimen is a flat-dose drug taken once daily, every day of the
medication cycle. Here a drug is dosed at 420 mg once daily on a continuous
28-day cycle (an ibrutinib-like schedule). Available tablet strengths are 140,
280, 420, and 560 mg, with a per-tablet cost derived from the pack price.

Start by defining the regimen with `define_regimen()`, then use
`summarise_regimen()` to inspect how the dose maps to tablets. The text summary
shows the tablet breakdown per administration, the dispensing mode (oral wastage
vs per-administration), the medication cycle schedule, and the treatment
duration cap.

```{r oral-continuous-define}
reg_continuous <- define_regimen(
  name = "Oral Drug A",
  route = "oral",
  dose_per_admin = 420,
  available_strengths = c(140, 280, 420, 560),
  unit_cost = 1430.80 / 28,
  med_cycle_length = 28,
  admin_days = 1:28
)

summarise_regimen(reg_continuous)
```

Now calculate cycle-by-cycle costs with `calculate_med_costs()`. The
`regimen_costs` table has one row per model cycle with the following columns:

- **Cycle** -- Sequential model cycle number.
- **Start Day** -- The first day of this model cycle (day 0 is the start of the
  model).
- **End Day** -- The day after the last day of this model cycle.
- **N Dispensing** -- Number of dispensing events that fall within this model
  cycle. For oral drugs with wastage (the default), one dispensing event
  occurs at the start of each medication cycle.
- **N Tablets** -- Total tablets dispensed in this model cycle (across all
  dispensing events).
- **Drug Cost** -- Total drug acquisition cost for this model cycle.
- **Admin Cost** -- Total administration/dispensing cost (here \$0).
- **Total Cost** -- Drug Cost + Admin Cost.

```{r oral-continuous-costs}
costs_continuous <- calculate_med_costs(
  reg_continuous,
  model_cycle_length = 28,
  n_cycles = 12
)

as_flextable(costs_continuous)
```

**Verify:** The optimal tablet combination for 420 mg is 1 x 420 mg tablet.
Each model cycle has 28 tablets at \$51.10 each, so Drug Cost = 28 x \$51.10
= \$1,430.80 per cycle.

### Oral wastage

By default, oral drugs use `oral_wastage = TRUE`: the full medication cycle's
supply is dispensed at the start of each medication cycle, and all drug cost is
assigned to the dispensing day. A patient who discontinues mid-cycle still
incurs the full dispensing cost. This models real-world pharmacy dispensing where
the entire cycle's tablets are handed to the patient upfront.

Setting `oral_wastage = FALSE` attributes cost per-administration instead, the
same way IV drugs are costed. Each administration day carries its own drug cost.

To see the difference, use `daily_detail = TRUE` for a row-per-day breakdown.
The daily detail table (oral wastage variant) has a two-row header with column
groups:

- **Medication Cycle** group:
  - *Med Cycle* -- Which medication cycle this day belongs to.
  - *Cycle Day* -- Day within the medication cycle (1-indexed).
- **Administration** group:
  - *Disp?* -- TRUE on dispensing days (day 1 of each medication cycle).
  - *Doses* -- Number of doses taken on this day (0 or 1).
  - *mg Taken* -- Milligrams actually taken on this day.
  - *mg Dispensed* -- Milligram content of the tablets taken on this day.
    This is a per-administration value, not the bulk cycle supply.
- **Dispensing (Costed)** group — these columns show the costing attribution:
  - *Doses Costed* -- Number of doses whose cost is assigned to this day.
    On a dispensing day, this equals the total administrations in the medication
    cycle. On non-dispensing days, this is 0.
  - *Tabs* -- Tablets taken on this day.
  - *Tabs Costed* -- Tablets whose cost is assigned to this day. On a dispensing
    day, this is the full cycle's worth of tablets.
  - *mg Costed* -- Milligrams whose cost is assigned to this day.
- **Costs** group:
  - *Drug Cost* -- Drug acquisition cost assigned to this day. On dispensing
    days, this is the full medication cycle's drug cost. On administration-only
    days, Drug Cost = \$0 because it was already charged at dispensing.
  - *Admin Cost*, *Total Cost*.

Here is a comparison — first with wastage (the default), where all cost is
front-loaded to the dispensing day:

```{r oral-wastage-daily}
daily_wastage <- calculate_med_costs(
  reg_continuous,
  model_cycle_length = 28,
  n_cycles = 2,
  daily_detail = TRUE
)
as_flextable(daily_wastage)
```

And the same regimen with `oral_wastage = FALSE`, where each administration day
carries its own drug cost:

```{r oral-no-wastage-daily}
reg_no_wastage <- define_regimen(
  name = "Oral Drug A (no wastage)",
  route = "oral",
  dose_per_admin = 420,
  available_strengths = c(140, 280, 420, 560),
  unit_cost = 1430.80 / 28,
  med_cycle_length = 28,
  admin_days = 1:28,
  oral_wastage = FALSE
)

daily_no_wastage <- calculate_med_costs(
  reg_no_wastage,
  model_cycle_length = 28,
  n_cycles = 2,
  daily_detail = TRUE
)
as_flextable(daily_no_wastage)
```

Note that the column schema differs between the two modes: with
`oral_wastage = TRUE`, the output contains `n_dispensing_events` and
`n_tablets_dispensed` columns (costing unit = dispensing event); with
`oral_wastage = FALSE`, it contains `n_administrations` (costing unit =
individual administration).

### Dispensing costs

The `admin_cost` parameter adds a dispensing or administration cost. For oral
drugs in the standard costing modes, `admin_cost` is charged once per dispensing
event (i.e., once per medication cycle, on day 1), regardless of the
`oral_wastage` setting. The `oral_wastage` parameter only controls how *drug*
cost is attributed — upfront at dispensing (`TRUE`) or spread across individual
administrations (`FALSE`). In `simple_daily_cost` mode, `admin_cost` is
distributed uniformly across the medication cycle (just like drug cost), so
`admin_cost / med_cycle_length` is charged per day.

Here is an example with a \$25 dispensing cost — the Admin Cost column shows a
single \$25 charge per cycle:

```{r oral-admin-cost}
reg_with_disp <- define_regimen(
  name = "Oral Drug A (with dispensing cost)",
  route = "oral",
  dose_per_admin = 420,
  available_strengths = c(140, 280, 420, 560),
  unit_cost = 1430.80 / 28,
  med_cycle_length = 28,
  admin_days = 1:28,
  admin_cost = 25,
  oral_wastage = TRUE
)

costs_with_disp <- calculate_med_costs(
  reg_with_disp,
  model_cycle_length = 28,
  n_cycles = 6
)

as_flextable(costs_with_disp)
```

**Verify:** Admin Cost = \$25.00 per cycle (one dispensing event per 28-day
medication cycle). Drug Cost is unchanged at \$1,430.80, so Total Cost =
\$1,455.80 per cycle. The same Admin Cost applies with `oral_wastage = FALSE`
— only the drug cost attribution changes, not the dispensing charge.

## 2. Costing an Oral Drug with Rest Periods

Many oral drugs are administered on only a subset of the medication cycle days.
Here a drug is dosed at 25 mg on days 1--21 of a 28-day cycle, with a 7-day
rest period (a lenalidomide-like schedule). The `admin_days` parameter defines
exactly which days within the medication cycle have an administration.

```{r oral-rest-define}
reg_rest <- define_regimen(
  name = "Oral Drug B",
  route = "oral",
  dose_per_admin = 25,
  available_strengths = c(2.5, 5, 7.5, 10, 15, 20, 25),
  unit_cost = 4368 / 21,
  med_cycle_length = 28,
  admin_days = 1:21,
  oral_wastage = TRUE
)

summarise_regimen(reg_rest)
```

The cost table shows 21 tablets dispensed per cycle (not 28), reflecting
the rest period:

```{r oral-rest-costs}
costs_rest <- calculate_med_costs(
  reg_rest,
  model_cycle_length = 28,
  n_cycles = 12
)

as_flextable(costs_rest)
```

**Verify:** Only 21 tablets are dispensed per medication cycle because
administration occurs only on days 1--21. Drug Cost = 21 x \$208.00 =
\$4,368.00 per cycle.

### Inspecting the event schedule with `detail_schedule()`

The `detail_schedule()` function produces a row-per-event view of the regimen,
showing every dispensing and administration event on the model timeline. This is
particularly useful for regimens with rest periods to verify which days are
active.

The `regimen_schedule` table has a two-row header with:

- **Standalone columns**: *Day* (absolute model day), *Model Cycle*, *Event
  Type* (either "dispensing" or "administration").
- **Medication Cycle** group: *Med Cycle*, *Cycle Day*.
- **Costs** group: *Drug Cost*, *Admin Cost*, *Total Cost*.

With oral wastage, dispensing events carry the drug cost; administration events
show the drug being taken but carry \$0 cost.

```{r oral-schedule}
schedule_rest <- detail_schedule(
  reg_rest,
  model_cycle_length = 28,
  n_cycles = 3
)
as_flextable(schedule_rest)
```

**Verify:** Each medication cycle produces 1 dispensing event and 21
administration events. Over 3 model cycles (= 3 medication cycles here,
since both are 28 days), there are 3 dispensing events and 63 administration
events.

## 3. Using a Flat Daily Cost

The `cost_per_day` parameter with `simple_daily_cost = TRUE` applies a flat
cost-per-mg rate that distributes costs uniformly over time, bypassing tablet
optimisation, wastage, and all dosing detail. No tablet breakdown, dispensing,
or wastage columns appear. This is useful when detailed dosing mechanics are
irrelevant and a simple per-day cost is sufficient.

```{r oral-simple}
reg_simple <- define_regimen(
  name = "Daily cost drug",
  route = "oral",
  cost_per_day = 5.00,
  admin_cost = 21,
  med_cycle_length = 21
)

sched_simple <- detail_schedule(
  reg_simple,
  model_cycle_length = 28,
  n_cycles = 3,
  simple_daily_cost = TRUE
)

as_flextable(sched_simple)
```

**Verify:** Every day shows the same \$5.00 drug cost and \$1.00 admin
cost (\$21.00 / 21 days), regardless of medication cycle boundaries.
The flat daily rate removes all dosing detail — no dispensing events, no
wastage, no tablet breakdown.

Now aggregate to model cycles:

```{r oral-simple-model}
costs_simple <- calculate_med_costs(
  reg_simple,
  model_cycle_length = 28,
  n_cycles = 12,
  simple_daily_cost = TRUE
)

as_flextable(costs_simple)
```

**Verify:** Each 28-day model cycle shows drug cost = 28 x \$5.00 =
\$140.00 and admin cost = 28 x \$1.00 = \$28.00, for a total of
\$168.00. The model cycle (28 days) differs from the medication cycle
(21 days) — the daily rate smooths costs uniformly regardless of cycle
alignment.

## 4. Capping Treatment Duration

The `max_med_cycles` parameter limits how many medication cycles are delivered,
even if the model horizon extends further. Here we limit treatment to 12
medication cycles over a 24-cycle model horizon. Costs drop to zero
after 12 x 28 = 336 days:

```{r oral-cap}
reg_capped <- define_regimen(
  name = "Oral Drug A (capped)",
  route = "oral",
  dose_per_admin = 420,
  available_strengths = c(140, 280, 420, 560),
  unit_cost = 1430.80 / 28,
  med_cycle_length = 28,
  admin_days = 1:28,
  max_med_cycles = 12
)

costs_capped <- calculate_med_costs(
  reg_capped,
  model_cycle_length = 28,
  n_cycles = 24
)

as_flextable(costs_capped)
```

**Verify:** Cycles 1--12 each show \$1,430.80 in drug cost. From cycle 13
onward (day 336+), all costs drop to \$0.00.

## 5. Offsetting Medication Start with `start_day`

The `start_day` parameter sets the day on which the first medication
cycle begins on the shared model timeline. Medication cycle 1 starts on
`start_day`, cycle 2 starts on `start_day + med_cycle_length`, and so on
(`start_day + (k-1) * med_cycle_length` for cycle *k*). Any model time
before `start_day` simply has no medication activity.

This is primarily useful for **chaining regimens** — for example, a
maintenance drug that begins after an induction period. By setting
`start_day = 56`, you position the entire medication timeline to start on
day 56, rather than day 0.

Here Oral Drug A starts on day 56 (after a 56-day induction phase),
with treatment capped at 6 medication cycles:

```{r oral-delayed-define}
reg_delayed <- define_regimen(
  name = "Oral Drug A (delayed)",
  route = "oral",
  dose_per_admin = 420,
  available_strengths = c(140, 280, 420, 560),
  unit_cost = 1430.80 / 28,
  med_cycle_length = 28,
  admin_days = 1:28,
  start_day = 56,
  max_med_cycles = 6
)
```

First, inspect the event-level schedule with `detail_schedule()`. The
**Med Cycle** column confirms that medication cycle 1 starts on day 56:

```{r oral-delayed-schedule}
sched_delayed <- detail_schedule(
  reg_delayed,
  model_cycle_length = 28,
  n_cycles = 10
)

as_flextable(sched_delayed)
```

**Verify:** Medication cycle 1 begins on day 56. All events fall within
medication cycles 1--6.

Now aggregate to **model cycles** with `calculate_med_costs()`. This is
how the costs integrate into a Markov model:

```{r oral-delayed-model-cycles}
costs_model_cycles <- calculate_med_costs(
  reg_delayed,
  model_cycle_length = 28,
  n_cycles = 10
)

as_flextable(costs_model_cycles)
```

**Verify:** Model cycles 1--2 (days 0--55) show zero costs because they
precede `start_day`. Medication cycle 1 maps to model cycle 3 (day 56),
so model cycles 3--8 each show \$1,430.80. Model cycles 9--10 are
beyond the 6-cycle cap and show \$0.00.
